---
title: "Market Sedimentation in Specialty Coffee"
author: "Guil Botler"
format:
  dashboard:
    orientation: rows
    scrolling: true
execute:
  echo: false
---

```{r}
library(tidyverse)
library(plotly)
library(DT)
library(scales)

data_path <- "data/processed/auction_lots.csv"
lots <- readr::read_csv(data_path, show_col_types = FALSE)
has_data <- nrow(lots) > 0 && any(!is.na(lots$final_bid_usd_lb))
entries <- lots |> distinct(entry_id, .keep_all = TRUE)
reported <- lots |> filter(auction_result_status == "reported")
```

## Overview

::: {.valuebox icon="cup-hot" color="primary"}
### Auction lots
```{r}
cat(if (has_data) comma(nrow(lots)) else "Awaiting validated extraction")
```
:::

::: {.valuebox icon="layers" color="secondary"}
### Coffee entries
```{r}
cat(if (has_data) comma(nrow(entries)) else "—")
```
:::

::: {.valuebox icon="currency-dollar" color="success"}
### Reported bids
```{r}
cat(if (has_data) comma(nrow(reported)) else "—")
```
:::

::: {.card}
### Scope

This dashboard measures the emergence and persistence of differentiated signals inside COE/NW auctions. Because public auction data contain selected finalists, the dashboard does not equate this sample with the whole coffee market or claim by itself to identify escape from a low-quality pooling equilibrium.

The unit distinction is explicit: an **auction lot** is a sale unit; a **coffee entry** is the underlying competition entry. Split lots such as 1A/1B share one `entry_id` and count once in composition measures.
:::

## Price signals

::: {.card}
### Score and final bid
```{r}
if (has_data) {
  plot_ly(
    lots |> filter(!is.na(score), !is.na(final_bid_usd_lb)),
    x = ~score, y = ~final_bid_usd_lb, color = ~program, symbol = ~process_group,
    text = ~paste(farm, process_group, variety, process, sep = "<br>"),
    type = "scatter", mode = "markers"
  ) |>
    layout(xaxis = list(title = "Jury score"), yaxis = list(title = "Final bid (USD/lb)", type = "log"))
} else {
  cat("No validated auction observations yet. Run `targets::tar_make()` locally.")
}
```
:::

::: {.card}
### Pilot price profile
```{r}
if (has_data) {
  lots |>
    group_by(program) |>
    summarise(
      `Auction lots` = n(),
      `Coffee entries` = n_distinct(entry_id),
      `Reported bids` = sum(!is.na(final_bid_usd_lb)),
      `Median bid (USD/lb)` = median(final_bid_usd_lb, na.rm = TRUE),
      `Minimum bid` = min(final_bid_usd_lb, na.rm = TRUE),
      `Maximum bid` = max(final_bid_usd_lb, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(across(where(is.numeric), ~round(.x, 2))) |>
    datatable(options = list(dom = "t"), rownames = FALSE)
}
```

Brazil 2024 is descriptive evidence from one event. The raw COE–NW gap combines score selection, program rules and market valuation; it is not yet a causal premium.
:::

::: {.card}
### Conditional interpretation

Price dispersion is not automatically differentiation. The empirical models first condition on score, auction, lot size and scarcity; only then do process, variety, origin and buyer matching enter as candidate organizing signals.
:::

## Differentiation and innovation

::: {.card}
### Process composition
```{r}
if (has_data) {
  composition <- entries |>
    count(year, process_group) |>
    group_by(year) |>
    mutate(share = n / sum(n)) |>
    ungroup()
  plot_ly(composition, x = ~year, y = ~share, color = ~process_group, type = "bar") |>
    layout(barmode = "stack", yaxis = list(title = "Share of coffee entries", tickformat = ".0%"))
} else {
  cat("The pilot will begin with Brazil 2024: Washed + Honey, Natural and Experimental.")
}
```
:::

::: {.card}
### Sedimentation scorecard

The dashboard will report breadth, recognition, price separation, buyer matching, reproduction and innovation persistence separately. A new label is not called innovation merely because it appears once.
:::

## Buyer matching

::: {.card}
### Buyer–attribute table
```{r}
if (has_data) {
  lots |>
    filter(!is.na(buyer), buyer != "") |>
    count(buyer, process_group, sort = TRUE) |>
    datatable(options = list(pageLength = 12), rownames = FALSE)
} else {
  cat("Buyer matching becomes available after auction and competition tables are joined and validated.")
}
```
:::

## Data and methods

::: {.card}
### Analytical data
```{r}
datatable(lots, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)
```
:::

::: {.card}
### Interpretation rules

- International coffee prices are used as benchmarks, not as the main instrumental variable.
- COE and NW are analyzed separately before pooling.
- Missing auction outcomes are never coded automatically as zero.
- Raw labels are preserved alongside later canonical taxonomies.
- Descriptive coefficients are not presented as causal effects.
:::
