---
title: "Market Sedimentation in Specialty Coffee"
author: "Guil Botler"
format:
  dashboard:
    orientation: rows
    scrolling: true
execute:
  echo: false
---

```{r}
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(crosstalk)

read_optional <- function(path) {
  if (!file.exists(path)) return(tibble())
  readr::read_csv(path, show_col_types = FALSE)
}

competition <- read_optional("data/processed/v2/competition_entries.csv")
auctions <- read_optional("data/processed/v2/auction_lots.csv")
benchmarks <- read_optional("data/processed/commodity_prices.csv")
coverage <- read_optional("data/processed/v2/process_coverage.csv")
downloads <- read_optional("data/audit/v2/download.csv")
reductions <- read_optional("data/audit/v2/entry_reduction.csv")

has_competition <- nrow(competition) > 0
has_auctions <- nrow(auctions) > 0

competition <- competition |>
  mutate(classification_group = case_when(
    process_classification_source %in% c("reported process", "ACE table heading") ~
      "Entry or table reported",
    process_classification_source == "ACE event title" ~ "Event-restricted category",
    TRUE ~ "Not reported"
  ))

auctions <- auctions |>
  mutate(entry_rank = if_else(
    program == "COE" & str_detect(rank_key, "^[0-9]+[AB]$"),
    str_remove(rank_key, "[AB]$"), rank_key
  ))

entry_attributes <- competition |>
  select(event_id, program, process_group, entry_rank, entry_key,
         competition_score = score, competition_farm = farm, producer,
         process, process_family, process_classification_source,
         classification_group, competition_variety = variety, region) |>
  group_by(event_id, program, process_group, entry_rank) |>
  filter(n() == 1L) |>
  ungroup()

matched_lots <- auctions |>
  left_join(entry_attributes,
            by = c("event_id", "program", "process_group", "entry_rank")) |>
  mutate(
    score_display = coalesce(competition_score, score),
    farm_display = coalesce(competition_farm, farm),
    variety_display = coalesce(competition_variety, variety),
    process_family = replace_na(process_family, "Not reported"),
    match_status = if_else(is.na(entry_key), "Unmatched or ambiguous", "Unique match")
  )

plot_config <- function(p, filename) plotly::config(
  p, displaylogo = FALSE, responsive = TRUE,
  toImageButtonOptions = list(format = "png", filename = filename)
)

safe_stat <- function(x, fun, min_n = 1L) {
  if (sum(!is.na(x)) < min_n) return(NA_real_)
  fun(x, na.rm = TRUE)
}
```

## V2 overview

::: {.valuebox icon="layers" color="primary"}
### Competition entries
```{r}
cat(if (has_competition) comma(nrow(competition)) else "Awaiting V2 data")
```
:::

::: {.valuebox icon="cup-hot" color="secondary"}
### Auction lots
```{r}
cat(if (has_auctions) comma(nrow(auctions)) else "—")
```
:::

::: {.valuebox icon="currency-dollar" color="success"}
### Reported bids
```{r}
cat(if (has_auctions) comma(sum(!is.na(auctions$final_bid_usd_lb))) else "—")
```
:::

::: {.valuebox icon="calendar3" color="warning"}
### Events
```{r}
cat(comma(n_distinct(c(competition$event_id, auctions$event_id))))
```
:::

::: {.valuebox icon="globe-americas" color="info"}
### Origins
```{r}
cat(comma(n_distinct(c(competition$country, auctions$country))))
```
:::

::: {.valuebox icon="calendar-range" color="dark"}
### Coverage
```{r}
all_years <- c(competition$year, auctions$year)
cat(paste(range(all_years, na.rm = TRUE), collapse = "–"))
```
:::

::: {.card}
### Scope

A **competition entry** is a selected coffee published by a COE/NW competition. An **auction lot** is a sale unit; one entry may be split into 1A/1B lots. Participation measures presence inside this selected specialty institution—not all submissions or national specialty production. Prices extend back to 1999, while named buyers and entry-level process labels have stronger recent coverage. Missing outcomes are never coded as zero.

Score, process and variety are added to a sale lot only when its event/program/process/rank key identifies one unique competition entry. Ambiguous keys remain unmatched rather than duplicating lots.
:::

## Participation

::: {.card}
### Published entries, auction lots and events
```{r}
annual <- bind_rows(
  competition |> count(year, name = "value") |>
    mutate(measure = "Competition entries"),
  competition |> distinct(year, event_id) |> count(year, name = "value") |>
    mutate(measure = "Competition events"),
  auctions |> count(year, name = "value") |>
    mutate(measure = "Auction lots"),
  auctions |> distinct(year, event_id) |> count(year, name = "value") |>
    mutate(measure = "Auction events")
)

plot_ly(annual, x = ~year, y = ~value, color = ~measure,
        type = "scatter", mode = "lines+markers",
        text = ~paste(measure, year, comma(value), sep = "<br>")) |>
  layout(xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Published observations"),
         legend = list(orientation = "h")) |>
  plot_config("annual-participation")
```

Growth can reflect participation, additional country programs, finalist limits, event design or publication coverage. It is an observable first-stage proxy, not yet a national entry rate.
:::

::: {.card}
### Country–year explorer
```{r}
country_year <- competition |>
  group_by(country, year, program) |>
  summarise(events = n_distinct(event_id), entries = n(),
            farms = n_distinct(farm_key[farm_key != "" & !is.na(farm_key)]),
            process_coverage = mean(process_classification_source != "not reported"),
            .groups = "drop") |>
  mutate(key = paste(country, year, program, sep = "|"))

shared_participation <- SharedData$new(country_year, key = ~key,
                                       group = "participation")
controls <- htmltools::div(
  filter_select("part_country", "Origins", shared_participation, ~country,
                multiple = TRUE),
  filter_slider("part_year", "Years", shared_participation, ~year, step = 1),
  filter_select("part_program", "Programs", shared_participation, ~program,
                multiple = TRUE)
)
p <- plot_ly(shared_participation, x = ~year, y = ~entries, color = ~country,
             linetype = ~program, type = "scatter", mode = "lines+markers",
             text = ~paste(country, program, paste0("Entries: ", entries),
                           paste0("Events: ", events),
                           paste0("Process coverage: ", percent(process_coverage)),
                           sep = "<br>")) |>
  layout(xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Published competition entries"),
         legend = list(orientation = "h")) |>
  plot_config("country-year-participation")
bscols(widths = c(3, 9), controls, p)
```
:::

## Process specialization

::: {.card}
### Composition with classification provenance
```{r}
process_data <- competition |>
  count(country, year, program, classification_group, process_family,
        name = "entries") |>
  group_by(country, year, program, classification_group) |>
  mutate(share = entries / sum(entries),
         key = paste(country, year, program, classification_group,
                     process_family, sep = "|")) |>
  ungroup()

shared_process <- SharedData$new(process_data, key = ~key, group = "process")
process_controls <- htmltools::div(
  filter_select("proc_country", "Origins", shared_process, ~country,
                multiple = TRUE),
  filter_slider("proc_year", "Years", shared_process, ~year, step = 1),
  filter_select("proc_program", "Programs", shared_process, ~program,
                multiple = TRUE),
  filter_select("proc_source", "Classification provenance", shared_process,
                ~classification_group, multiple = TRUE)
)
pp <- plot_ly(shared_process, x = ~year, y = ~share, color = ~process_family,
              linetype = ~country, type = "scatter", mode = "lines+markers",
              text = ~paste(country, program, classification_group,
                            process_family, paste0("Entries: ", entries),
                            percent(share, accuracy = 0.1), sep = "<br>")) |>
  layout(xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Share within provenance group", tickformat = ".0%"),
         legend = list(orientation = "h")) |>
  plot_config("process-specialization")
bscols(widths = c(3, 9), process_controls, pp)
```

Use **Entry or table reported** for the main specialization analysis. **Event-restricted category** identifies explicit Natural or Pulped Natural events, mainly in Brazil; it documents institutional categories, not producer-choice shares from an open multi-process competition.
:::

::: {.card}
### Process information coverage
```{r}
source_year <- competition |> count(year, classification_group, name = "entries") |>
  group_by(year) |> mutate(share = entries / sum(entries)) |> ungroup()
plot_ly(source_year, x = ~year, y = ~share, color = ~classification_group,
        type = "bar", text = ~paste(classification_group,
                                     paste0("Entries: ", entries),
                                     percent(share, accuracy = 0.1), sep = "<br>")) |>
  layout(barmode = "stack", xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Share of entries", tickformat = ".0%"),
         legend = list(orientation = "h")) |>
  plot_config("process-coverage")
```
:::

## Prices and differentiated-market benchmark

::: {.card}
### Score and final bid—matched V2 observations
```{r}
priced <- matched_lots |>
  filter(!is.na(final_bid_usd_lb), !is.na(score_display)) |>
  mutate(key = lot_key)
shared_prices <- SharedData$new(priced, key = ~key, group = "prices")
price_controls <- htmltools::div(
  filter_select("price_country", "Origins", shared_prices, ~country,
                multiple = TRUE),
  filter_slider("price_year", "Years", shared_prices, ~year, step = 1),
  filter_select("price_program", "Programs", shared_prices, ~program,
                multiple = TRUE),
  filter_select("price_process", "Processes", shared_prices, ~process_family,
                multiple = TRUE)
)
sp <- plot_ly(shared_prices, x = ~score_display, y = ~final_bid_usd_lb,
              color = ~country, symbol = ~process_family,
              type = "scatter", mode = "markers",
              text = ~paste(country, year, farm_display, process_family,
                            variety_display, paste0("Score: ", score_display),
                            paste0("USD/lb: ", final_bid_usd_lb), sep = "<br>")) |>
  layout(xaxis = list(title = "Jury score"),
         yaxis = list(title = "Final bid (USD/lb)", type = "log"),
         legend = list(orientation = "h")) |>
  plot_config("score-final-bid")
bscols(widths = c(3, 9), price_controls, sp)
```
:::

::: {.card}
### Auction statistics
```{r}
auctions |> group_by(country, year, program) |>
  summarise(`Auction lots` = n(),
            `Reported bids` = sum(!is.na(final_bid_usd_lb)),
            `Mean bid` = safe_stat(final_bid_usd_lb, mean),
            `Variance` = safe_stat(final_bid_usd_lb, var, 2L),
            `Median bid` = safe_stat(final_bid_usd_lb, median),
            `Minimum` = safe_stat(final_bid_usd_lb, min),
            `Maximum` = safe_stat(final_bid_usd_lb, max), .groups = "drop") |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  datatable(options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE)
```
:::

::: {.card}
### Median auction bid relative to Arabica
```{r}
if (nrow(benchmarks) > 0) {
  arabica <- benchmarks |> filter(series == "Coffee Arabica — World Bank") |>
    group_by(year) |>
    summarise(arabica_usd_lb = mean(value, na.rm = TRUE) / 2.2046226218,
              .groups = "drop")
  ratio <- auctions |> filter(!is.na(final_bid_usd_lb)) |>
    group_by(country, year, program) |>
    summarise(median_bid = median(final_bid_usd_lb), priced_lots = n(),
              .groups = "drop") |>
    left_join(arabica, by = "year") |>
    mutate(ratio = median_bid / arabica_usd_lb)
  plot_ly(ratio, x = ~year, y = ~ratio, color = ~country,
          linetype = ~program, type = "scatter", mode = "lines+markers",
          text = ~paste(country, program, paste0("Lots: ", priced_lots),
                        paste0(round(ratio, 1), "×"), sep = "<br>")) |>
    layout(xaxis = list(title = "Auction year", dtick = 1),
           yaxis = list(title = "Median auction bid / Arabica benchmark"),
           legend = list(orientation = "h")) |>
    plot_config("specialty-benchmark-ratio")
}
```

This ratio is a descriptive differentiated-market premium—not a causal effect and not a farm-gate margin.
:::

## Buyer depth

::: {.card}
### Named-buyer coverage
```{r}
buyer_year <- auctions |> group_by(year) |>
  summarise(lots = n(), buyer_reported = sum(!is.na(buyer) & buyer != ""),
            named_buyers = n_distinct(buyer[!is.na(buyer) & buyer != ""]),
            coverage = buyer_reported / lots, .groups = "drop")
plot_ly(buyer_year, x = ~year, y = ~coverage, type = "bar",
        text = ~paste(year, paste0("Named-buyer lots: ", buyer_reported),
                      paste0("Distinct buyers: ", named_buyers),
                      percent(coverage, accuracy = 0.1), sep = "<br>")) |>
  layout(xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Share with named buyer", tickformat = ".0%")) |>
  plot_config("buyer-coverage")
```

Buyer analysis is defensible mainly from 2018 onward. Earlier missing names mean “not published or not recognized,” not “no buyer.”
:::

## Coverage and audit

::: {.card}
### Event-level process coverage
```{r}
datatable(coverage |> mutate(coverage_rate = percent(coverage_rate, accuracy = 0.1)),
          options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE,
          filter = "top")
```
:::

::: {.card}
### Download audit
```{r}
datatable(downloads, options = list(pageLength = 15, scrollX = TRUE),
          rownames = FALSE, filter = "top")
```
:::

::: {.card}
### Raw-table reduction audit
```{r}
datatable(reductions |> arrange(desc(abs(removed))),
          options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE,
          filter = "top")
```

Three fully overlapping sparse summary tables were excluded before harmonization. Remaining reductions still require classification before freezing the ACE analytical base.
:::

::: {.card}
### Interpretation rules

- Entries and auction lots are separate units; split lots never duplicate process choices.
- Process provenance remains explicit: reported, event-restricted or missing.
- Historical prices cover a longer period than buyers and process labels.
- COE and NW remain separately filterable.
- Participation means published finalists, not national adoption.
- Descriptive price ratios and coefficients are not causal effects.
:::

::: {.card}
### Data sources

- [Alliance for Coffee Excellence historical results](https://allianceforcoffeeexcellence.org/competition-auction-results/)
- [World Bank Commodity Price Data](https://www.worldbank.org/en/research/commodity-markets)
- [BLS winery producer price index via FRED](https://fred.stlouisfed.org/series/PCU3121303121300)
:::
