---
title: "Market Sedimentation in Specialty Coffee"
author: "Guil Botler"
format:
  dashboard:
    orientation: rows
    scrolling: true
execute:
  echo: false
---

```{r}
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(crosstalk)

read_optional_csv <- function(path) {
  if (!file.exists(path)) return(tibble())
  readr::read_csv(path, show_col_types = FALSE)
}

lots <- read_optional_csv("data/processed/auction_lots.csv")
benchmarks <- read_optional_csv("data/processed/commodity_prices.csv")
source_audit <- read_optional_csv("data/processed/source_audit.csv")

has_data <- nrow(lots) > 0 && any(!is.na(lots$final_bid_usd_lb))
if (nrow(lots) > 0 && !"process_family" %in% names(lots)) {
  lots <- lots |> mutate(process_family = coalesce(process_group, "Not reported"))
}
entries <- lots |> distinct(entry_id, .keep_all = TRUE)
reported <- lots |> filter(auction_result_status == "reported")

safe_stat <- function(x, fun, min_n = 1L) {
  if (sum(!is.na(x)) < min_n) return(NA_real_)
  fun(x, na.rm = TRUE)
}

apply_plot_config <- function(plot) {
  plotly::config(
    plot,
    displaylogo = FALSE,
    responsive = TRUE,
    toImageButtonOptions = list(format = "png", filename = "market-sedimentation")
  )
}
```

## Overview

::: {.valuebox icon="cup-hot" color="primary"}
### Auction lots
```{r}
cat(if (has_data) comma(nrow(lots)) else "Awaiting validated extraction")
```
:::

::: {.valuebox icon="layers" color="secondary"}
### Coffee entries
```{r}
cat(if (has_data) comma(nrow(entries)) else "—")
```
:::

::: {.valuebox icon="currency-dollar" color="success"}
### Reported bids
```{r}
cat(if (has_data) comma(nrow(reported)) else "—")
```
:::

::: {.valuebox icon="globe-americas" color="info"}
### Producing countries
```{r}
cat(if (has_data) comma(n_distinct(lots$country)) else "—")
```
:::

::: {.valuebox icon="calendar3" color="warning"}
### Auction years
```{r}
cat(if (has_data) paste(range(lots$year), collapse = "–") else "—")
```
:::

::: {.card}
### Scope

This dashboard measures the emergence and persistence of differentiated signals inside COE/NW auctions. Public auction data contain selected finalists, so this sample is not the whole coffee market and does not by itself identify escape from a low-quality pooling equilibrium.

An **auction lot** is a sale unit; a **coffee entry** is the underlying competition entry. Split lots such as 1A/1B share one `entry_id` and count once in composition measures. Missing auction outcomes are never coded as zero.
:::

## Interactive auction explorer

::: {.card}
### Choose origins, years, program and process
```{r}
if (has_data) {
  shared_lots <- SharedData$new(lots, key = ~lot_id, group = "auction-explorer")
  controls <- div(
    filter_select(
      "country_filter", "Producing origins", shared_lots, ~country,
      multiple = TRUE
    ),
    filter_slider(
      "year_filter", "Auction years", shared_lots, ~year,
      step = 1
    ),
    filter_select(
      "program_filter", "Programs", shared_lots, ~program,
      multiple = TRUE
    ),
    filter_select(
      "process_filter", "Canonical process families", shared_lots, ~process_family,
      multiple = TRUE
    )
  )
  bscols(widths = c(3, 9), controls, div(
    "Filters in this panel affect the scatter plot and lot table below. ",
    "Use the Plotly toolbar to zoom, select and download a PNG."
  ))
} else {
  cat("No validated auction observations yet.")
}
```
:::

::: {.card}
### Score and final bid
```{r}
if (has_data) {
  plot_ly(
    shared_lots,
    x = ~score, y = ~final_bid_usd_lb, color = ~program,
    symbol = ~process_family,
    text = ~paste(
      country, year, farm, process_family, variety,
      paste0("USD/lb: ", final_bid_usd_lb),
      sep = "<br>"
    ),
    type = "scatter", mode = "markers"
  ) |>
    layout(
      xaxis = list(title = "Jury score"),
      yaxis = list(title = "Final bid (USD/lb)", type = "log"),
      legend = list(orientation = "h")
    ) |>
    apply_plot_config()
}
```
:::

::: {.card}
### Filtered auction lots
```{r}
if (has_data) {
  datatable(
    shared_lots,
    options = list(pageLength = 12, scrollX = TRUE),
    rownames = FALSE,
    filter = "top"
  )
}
```
:::

## Process evolution

::: {.card}
### Annual participation of process families by country
```{r}
if (has_data) {
  process_composition <- entries |>
    count(country, year, program, process_family, name = "entries") |>
    group_by(country, year, program) |>
    mutate(
      total_entries = sum(entries),
      share = entries / total_entries,
      process_key = paste(country, year, program, process_family, sep = "|")
    ) |>
    ungroup()

  shared_process <- SharedData$new(
    process_composition,
    key = ~process_key,
    group = "process-evolution"
  )

  process_controls <- div(
    filter_select(
      "process_country_filter", "Producing origins", shared_process, ~country,
      multiple = TRUE
    ),
    filter_slider(
      "process_year_filter", "Auction years", shared_process, ~year,
      step = 1
    ),
    filter_select(
      "process_program_filter", "Programs", shared_process, ~program,
      multiple = TRUE
    )
  )

  process_plot <- plot_ly(
    shared_process,
    x = ~year, y = ~share, color = ~process_family,
    linetype = ~country,
    text = ~paste(
      country, program, process_family,
      paste0("Entries: ", entries),
      paste0("Share: ", percent(share, accuracy = 0.1)),
      sep = "<br>"
    ),
    type = "scatter", mode = "lines+markers"
  ) |>
    layout(
      xaxis = list(title = "Auction year", dtick = 1),
      yaxis = list(title = "Share of competition entries", tickformat = ".0%"),
      legend = list(orientation = "h")
    ) |>
    apply_plot_config()

  bscols(widths = c(3, 9), process_controls, process_plot)
} else {
  cat("No validated competition composition is available yet.")
}
```

The denominator is the number of distinct competition entries in each country–year–program. The canonical taxonomy preserves the reported process and maps it into: Washed, Natural, Honey/pulped natural, Experimental, grouped Natural/Honey, or Not reported. A missing process remains visible in the denominator.
:::

::: {.card}
### Process composition table
```{r}
if (has_data) {
  datatable(
    shared_process,
    options = list(pageLength = 12, scrollX = TRUE),
    rownames = FALSE
  )
}
```
:::

## Prices and external benchmarks

::: {.card}
### Auction bid statistics by event and program
```{r}
if (has_data) {
  lots |>
    group_by(country, year, program) |>
    summarise(
      `Auction lots` = n(),
      `Coffee entries` = n_distinct(entry_id),
      `Reported bids` = sum(!is.na(final_bid_usd_lb)),
      `Mean bid (USD/lb)` = safe_stat(final_bid_usd_lb, mean),
      `Bid variance` = safe_stat(final_bid_usd_lb, var, min_n = 2L),
      `Median bid (USD/lb)` = safe_stat(final_bid_usd_lb, median),
      `Minimum bid` = safe_stat(final_bid_usd_lb, min),
      `Maximum bid` = safe_stat(final_bid_usd_lb, max),
      .groups = "drop"
    ) |>
    mutate(across(where(is.numeric), ~round(.x, 2))) |>
    datatable(options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE)
}
```

These summaries are descriptive. Differences across countries, years, programs and processes combine score selection, auction rules, sample composition and market valuation; they are not causal premiums.
:::

::: {.card}
### Benchmark price evolution (common index)
```{r}
if (nrow(benchmarks) > 0) {
  benchmark_index <- benchmarks |>
    group_by(series) |>
    mutate(
      base_value = mean(value[year == 2020], na.rm = TRUE),
      base_value = if_else(is.nan(base_value), first(value[!is.na(value)]), base_value),
      index_2020 = 100 * value / base_value
    ) |>
    ungroup()

  plot_ly(
    benchmark_index,
    x = ~date, y = ~index_2020, color = ~series,
    text = ~paste(series, date, paste0("Index: ", round(index_2020, 1)), sep = "<br>"),
    type = "scatter", mode = "lines"
  ) |>
    layout(
      xaxis = list(title = "", rangeslider = list(visible = TRUE)),
      yaxis = list(title = "Index (2020 average = 100)"),
      legend = list(orientation = "h")
    ) |>
    apply_plot_config()
} else {
  cat("Benchmark data will appear after the validated refresh workflow finishes.")
}
```

Arabica, robusta and cocoa are monthly World Bank commodity prices in USD/kg. Wine is not a homogeneous global commodity: the wine comparison is the monthly U.S. winery producer price index from BLS/FRED, clearly kept as an index rather than converted into a fictitious global spot price.
:::

::: {.card}
### Benchmark levels and units
```{r}
if (nrow(benchmarks) > 0) {
  benchmarks |>
    group_by(series, unit, market_scope) |>
    filter(date == max(date, na.rm = TRUE)) |>
    ungroup() |>
    transmute(
      Series = series,
      Date = date,
      Value = round(value, 3),
      Unit = unit,
      Scope = market_scope
    ) |>
    datatable(options = list(dom = "t", scrollX = TRUE), rownames = FALSE)
}
```
:::

::: {.card}
### Specialty auction price relative to Arabica benchmark
```{r}
if (has_data && nrow(benchmarks) > 0) {
  arabica_annual <- benchmarks |>
    filter(series == "Coffee Arabica — World Bank") |>
    group_by(year) |>
    summarise(
      arabica_usd_lb = mean(value, na.rm = TRUE) / 2.2046226218,
      .groups = "drop"
    )

  auction_ratio <- lots |>
    filter(!is.na(final_bid_usd_lb)) |>
    group_by(country, year, program) |>
    summarise(median_auction_usd_lb = median(final_bid_usd_lb), .groups = "drop") |>
    left_join(arabica_annual, by = "year") |>
    mutate(ratio = median_auction_usd_lb / arabica_usd_lb)

  plot_ly(
    auction_ratio,
    x = ~year, y = ~ratio, color = ~country, linetype = ~program,
    text = ~paste(country, program, paste0(round(ratio, 1), "×"), sep = "<br>"),
    type = "scatter", mode = "lines+markers"
  ) |>
    layout(
      xaxis = list(title = "Auction year", dtick = 1),
      yaxis = list(title = "Median auction bid / Arabica benchmark"),
      legend = list(orientation = "h")
    ) |>
    apply_plot_config()
}
```

This ratio is a descriptive benchmark, not a causal premium and not a farm-gate margin.
:::

## Coverage and methods

::: {.card}
### Source coverage audit
```{r}
if (nrow(source_audit) > 0) {
  datatable(
    source_audit,
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE,
    filter = "top"
  )
} else {
  cat("The source audit will appear after the expanded registry is processed.")
}
```

Every registered country–year is either included or retained in this audit with its exclusion reason. A page that exists but fails the competition–auction join is not silently admitted.
:::

::: {.card}
### Interpretation rules

- International coffee prices are benchmarks, not the main instrumental variable.
- COE and NW are analyzed separately before pooling.
- Missing auction outcomes are never coded automatically as zero.
- Raw process labels and ACE headings are preserved alongside the canonical taxonomy.
- The process chart counts competition entries, not split auction lots.
- Descriptive coefficients are not presented as causal effects.
:::

::: {.card}
### Data sources

- Auction and competition tables: [Alliance for Coffee Excellence historical results](https://allianceforcoffeeexcellence.org/competition-auction-results/)
- Arabica, robusta and cocoa: [World Bank Commodity Price Data (Pink Sheet)](https://www.worldbank.org/en/research/commodity-markets)
- Wine comparison: [BLS winery producer price index via FRED](https://fred.stlouisfed.org/series/PCU3121303121300)
:::
