---
title: "Market Sedimentation in Specialty Coffee"
author: "Guil Botler"
format:
  dashboard:
    orientation: rows
    scrolling: true
execute:
  echo: false
---

```{r}
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(crosstalk)

data_path <- "data/processed/auction_lots.csv"
lots <- readr::read_csv(data_path, show_col_types = FALSE)
commodity_path <- "data/processed/commodity_prices.csv"
audit_path <- "data/processed/source_audit.csv"
commodity_prices <- if (file.exists(commodity_path)) {
  readr::read_csv(commodity_path, show_col_types = FALSE) |>
    mutate(date = as.Date(date))
} else {
  tibble()
}
source_audit <- if (file.exists(audit_path)) {
  readr::read_csv(audit_path, show_col_types = FALSE)
} else {
  tibble()
}
has_data <- nrow(lots) > 0 && any(!is.na(lots$final_bid_usd_lb))
lots <- lots |>
  group_by(entry_id) |>
  mutate(
    entry_weight = 1 / n(),
    process_display = if_else(
      is.na(process_group) | process_group == "",
      "Not reported",
      process_group
    )
  ) |>
  ungroup()
entries <- lots |> distinct(entry_id, .keep_all = TRUE)
reported <- lots |> filter(auction_result_status == "reported")
shared_lots <- SharedData$new(lots, key = ~lot_id, group = "auction_filters")
shared_price_lots <- SharedData$new(
  lots |> filter(!is.na(score), !is.na(final_bid_usd_lb)),
  key = ~lot_id,
  group = "auction_filters"
)
safe_stat <- function(x, fun, min_n = 1L) {
  if (sum(!is.na(x)) < min_n) return(NA_real_)
  fun(x, na.rm = TRUE)
}
```

## Overview

::: {.valuebox icon="cup-hot" color="primary"}
### Auction lots
```{r}
cat(if (has_data) comma(nrow(lots)) else "Awaiting validated extraction")
```
:::

::: {.valuebox icon="layers" color="secondary"}
### Coffee entries
```{r}
cat(if (has_data) comma(nrow(entries)) else "—")
```
:::

::: {.valuebox icon="currency-dollar" color="success"}
### Reported bids
```{r}
cat(if (has_data) comma(nrow(reported)) else "—")
```
:::

::: {.valuebox icon="globe-americas" color="info"}
### Producing countries
```{r}
cat(if (has_data) comma(n_distinct(lots$country)) else "—")
```
:::

::: {.valuebox icon="calendar3" color="warning"}
### Auction years
```{r}
cat(if (has_data) comma(n_distinct(lots$year)) else "—")
```
:::

::: {.valuebox icon="collection" color="primary"}
### Included events
```{r}
cat(if (has_data) comma(n_distinct(paste(lots$country, lots$year))) else "—")
```
:::

::: {.card}
### Scope

This dashboard measures the emergence and persistence of differentiated signals inside COE/NW auctions. Because public auction data contain selected finalists, the dashboard does not equate this sample with the whole coffee market or claim by itself to identify escape from a low-quality pooling equilibrium.

The unit distinction is explicit: an **auction lot** is a sale unit; a **coffee entry** is the underlying competition entry. Split lots such as 1A/1B share one `entry_id` and count once in composition measures.
:::

::: {.card}
### Interactive filters

```{r}
if (has_data) {
  bscols(
    widths = c(3, 3, 3, 3),
    filter_select("origin_filter", "Producing origin", shared_lots, ~country, multiple = TRUE),
    filter_slider("year_filter", "Auction years", shared_lots, ~year, step = 1),
    filter_select("program_filter", "Program", shared_lots, ~program, multiple = TRUE),
    filter_select("process_filter", "Process group", shared_lots, ~process_display, multiple = TRUE)
  )
}
```

The filters are linked to the lot-level price chart and analytical table. Plotly controls allow zooming, panning, hiding series through the legend and downloading the current view as PNG.
:::

## Price signals

::: {.card}
### Score and final bid
```{r}
if (has_data) {
  plot_ly(
    shared_price_lots,
    x = ~score, y = ~final_bid_usd_lb, color = ~program, symbol = ~process_group,
    text = ~paste(farm, process_group, variety, process, sep = "<br>"),
    type = "scatter", mode = "markers"
  ) |>
    layout(xaxis = list(title = "Jury score"), yaxis = list(title = "Final bid (USD/lb)", type = "log")) |>
    config(
      displaylogo = FALSE,
      toImageButtonOptions = list(format = "png", filename = "auction_score_and_bid")
    )
} else {
  cat("No validated auction observations yet. Run `targets::tar_make()` locally.")
}
```
:::

::: {.card}
### Bid statistics by event and program
```{r}
if (has_data) {
  lots |>
    group_by(country, year, program) |>
    summarise(
      `Auction lots` = n(),
      `Coffee entries` = n_distinct(entry_id),
      `Reported bids` = sum(!is.na(final_bid_usd_lb)),
      `Mean bid (USD/lb)` = safe_stat(final_bid_usd_lb, mean),
      `Bid variance` = safe_stat(final_bid_usd_lb, var, min_n = 2L),
      `Median bid (USD/lb)` = safe_stat(final_bid_usd_lb, median),
      `Minimum bid` = safe_stat(final_bid_usd_lb, min),
      `Maximum bid` = safe_stat(final_bid_usd_lb, max),
      .groups = "drop"
    ) |>
    mutate(across(where(is.numeric), ~round(.x, 2))) |>
    datatable(options = list(dom = "t"), rownames = FALSE)
}
```

These event-level summaries are descriptive. Raw differences across countries, years, COE/NW programs and processes combine score selection, auction rules, sample composition and market valuation; they are not causal premiums.
:::

::: {.card}
### Bid statistics by event and process
```{r}
if (has_data) {
  lots |>
    group_by(country, year, program, process_group) |>
    summarise(
      `Reported bids` = sum(!is.na(final_bid_usd_lb)),
      `Mean (USD/lb)` = safe_stat(final_bid_usd_lb, mean),
      `Variance` = safe_stat(final_bid_usd_lb, var, min_n = 2L),
      `Median (USD/lb)` = safe_stat(final_bid_usd_lb, median),
      `Minimum` = safe_stat(final_bid_usd_lb, min),
      `Maximum` = safe_stat(final_bid_usd_lb, max),
      .groups = "drop"
    ) |>
    mutate(across(where(is.numeric), ~round(.x, 2))) |>
    datatable(options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)
}
```

Variance is the sample variance of observed final bids within each group. Missing auction outcomes are excluded rather than assigned a zero price.
:::

::: {.card}
### Conditional interpretation

Price dispersion is not automatically differentiation. The empirical models first condition on score, auction, lot size and scarcity; only then do process, variety, origin and buyer matching enter as candidate organizing signals.
:::

## Differentiation and innovation

::: {.card}
### Process composition
```{r}
if (has_data) {
  composition <- entries |>
    count(country, year, process_display) |>
    group_by(country, year) |>
    mutate(share = n / sum(n)) |>
    ungroup()
  plot_ly(
    composition,
    x = ~year,
    y = ~share,
    color = ~process_display,
    frame = ~country,
    text = ~paste0(
      "Origin: ", country,
      "<br>Year: ", year,
      "<br>Process: ", process_display,
      "<br>Entries: ", n,
      "<br>Share: ", percent(share, accuracy = 0.1)
    ),
    hoverinfo = "text",
    type = "bar"
  ) |>
    layout(
      barmode = "stack",
      xaxis = list(title = "Auction year", dtick = 1),
      yaxis = list(title = "Share of coffee entries", tickformat = ".0%")
    ) |>
    animation_opts(700, redraw = TRUE) |>
    animation_slider(currentvalue = list(prefix = "Origin: ")) |>
    config(
      displaylogo = FALSE,
      toImageButtonOptions = list(format = "png", filename = "process_share_by_origin_year")
    )
} else {
  cat("No validated competition composition is available yet.")
}
```
:::

::: {.card}
### Commodity benchmarks
```{r}
if (nrow(commodity_prices) > 0) {
  plot_ly(
    commodity_prices,
    x = ~date,
    y = ~price_usd_kg,
    color = ~commodity,
    type = "scatter",
    mode = "lines",
    hovertemplate = paste(
      "%{x|%b %Y}",
      "<br>%{fullData.name}",
      "<br>USD/kg: %{y:.2f}<extra></extra>"
    )
  ) |>
    layout(
      xaxis = list(title = "Month", rangeslider = list(visible = TRUE)),
      yaxis = list(title = "World Bank monthly price (USD/kg)"),
      legend = list(orientation = "h")
    ) |>
    config(
      displaylogo = FALSE,
      toImageButtonOptions = list(format = "png", filename = "commodity_benchmarks")
    )
} else {
  cat("Commodity series are awaiting the next validated data refresh.")
}
```

Monthly benchmarks: World Bank Pink Sheet for Coffee Arabica, Coffee Robusta and Cocoa. Wine is not included until a specific market definition is selected because there is no single global wine commodity price directly comparable to these series.
:::

::: {.card}
### Sedimentation scorecard

The dashboard will report breadth, recognition, price separation, buyer matching, reproduction and innovation persistence separately. A new label is not called innovation merely because it appears once.
:::

## Buyer matching

::: {.card}
### Buyer–attribute table
```{r}
if (has_data) {
  lots |>
    filter(!is.na(buyer), buyer != "") |>
    count(country, year, buyer, process_group, sort = TRUE) |>
    datatable(options = list(pageLength = 12), rownames = FALSE)
} else {
  cat("Buyer matching becomes available after auction and competition tables are joined and validated.")
}
```
:::

## Data and methods

::: {.card}
### Analytical data
```{r}
datatable(
  shared_lots,
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  filter = "top"
)
```
:::

::: {.card}
### Source coverage audit
```{r}
if (nrow(source_audit) > 0) {
  source_audit |>
    select(country, year, status, auction_lots, coffee_entries, reported_bids, message, source_url) |>
    datatable(options = list(pageLength = 12, scrollX = TRUE), rownames = FALSE, filter = "top")
} else {
  cat("The source audit will appear after the expanded collection runs.")
}
```
:::

::: {.card}
### Interpretation rules

- International coffee prices are used as benchmarks, not as the main instrumental variable.
- COE and NW are analyzed separately before pooling.
- Missing auction outcomes are never coded automatically as zero.
- Raw labels are preserved alongside later canonical taxonomies.
- Descriptive coefficients are not presented as causal effects.
:::
