---
title: "Market Sedimentation in Specialty Coffee"
author: "Guil Botler"
format:
  dashboard:
    orientation: rows
    scrolling: true
execute:
  echo: false
---

```{r}
library(tidyverse)
library(plotly)
library(DT)
library(scales)

data_path <- "data/processed/auction_lots.csv"
lots <- readr::read_csv(data_path, show_col_types = FALSE)
has_data <- nrow(lots) > 0 && any(!is.na(lots$final_bid_usd_lb))
```

## Overview

::: {.valuebox icon="cup-hot" color="primary"}
### Observed lots
```{r}
if (has_data) comma(nrow(lots)) else "Awaiting validated extraction"
```
:::

::: {.valuebox icon="globe-americas" color="secondary"}
### Producing countries
```{r}
if (has_data) n_distinct(lots$country) else "—"
```
:::

::: {.valuebox icon="calendar3" color="success"}
### Auction years
```{r}
if (has_data) paste(min(lots$year), max(lots$year), sep = "–") else "—"
```
:::

::: {.card}
### Scope

This dashboard measures the emergence and persistence of differentiated signals inside COE/NW auctions. Because public auction data contain selected finalists, the dashboard does not equate this sample with the whole coffee market or claim by itself to identify escape from a low-quality pooling equilibrium.
:::

## Price signals

::: {.card}
### Score and final bid
```{r}
if (has_data) {
  plot_ly(
    lots |> filter(!is.na(score), !is.na(final_bid_usd_lb)),
    x = ~score, y = ~final_bid_usd_lb, color = ~process_group,
    text = ~paste(farm, variety, process, sep = "<br>"),
    type = "scatter", mode = "markers"
  ) |>
    layout(xaxis = list(title = "Jury score"), yaxis = list(title = "Final bid (USD/lb)", type = "log"))
} else {
  cat("No validated auction observations yet. Run `targets::tar_make()` locally.")
}
```
:::

::: {.card}
### Conditional interpretation

Price dispersion is not automatically differentiation. The empirical models first condition on score, auction, lot size and scarcity; only then do process, variety, origin and buyer matching enter as candidate organizing signals.
:::

## Differentiation and innovation

::: {.card}
### Process composition
```{r}
if (has_data) {
  composition <- lots |>
    count(year, process_group) |>
    group_by(year) |>
    mutate(share = n / sum(n)) |>
    ungroup()
  plot_ly(composition, x = ~year, y = ~share, color = ~process_group, type = "bar") |>
    layout(barmode = "stack", yaxis = list(title = "Share of listed lots", tickformat = ".0%"))
} else {
  cat("The pilot will begin with Brazil 2024: Washed + Honey, Natural and Experimental.")
}
```
:::

::: {.card}
### Sedimentation scorecard

The dashboard will report breadth, recognition, price separation, buyer matching, reproduction and innovation persistence separately. A new label is not called innovation merely because it appears once.
:::

## Buyer matching

::: {.card}
### Buyer–attribute table
```{r}
if (has_data) {
  lots |>
    filter(!is.na(buyer), buyer != "") |>
    count(buyer, process_group, sort = TRUE) |>
    datatable(options = list(pageLength = 12), rownames = FALSE)
} else {
  cat("Buyer matching becomes available after auction and competition tables are joined and validated.")
}
```
:::

## Data and methods

::: {.card}
### Analytical data
```{r}
datatable(lots, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)
```
:::

::: {.card}
### Interpretation rules

- International coffee prices are used as benchmarks, not as the main instrumental variable.
- COE and NW are analyzed separately before pooling.
- Missing auction outcomes are never coded automatically as zero.
- Raw labels are preserved alongside later canonical taxonomies.
- Descriptive coefficients are not presented as causal effects.
:::

