---
title: "Market Sedimentation in Specialty Coffee"
author: "Guil Botler"
format:
  dashboard:
    orientation: rows
    scrolling: true
execute:
  echo: false
---

```{r}
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(crosstalk)

read_optional <- function(path) {
  if (!file.exists(path)) return(tibble())
  readr::read_csv(path, show_col_types = FALSE)
}

competition <- read_optional("data/processed/v2/competition_entries.csv")
auctions <- read_optional("data/processed/v2/auction_lots.csv")
benchmarks <- read_optional("data/processed/commodity_prices.csv")
coverage <- read_optional("data/processed/v2/process_coverage.csv")
downloads <- read_optional("data/audit/v2/download.csv")
reductions <- read_optional("data/audit/v2/entry_reduction.csv")

has_competition <- nrow(competition) > 0
has_auctions <- nrow(auctions) > 0

competition <- competition |>
  mutate(classification_group = case_when(
    process_classification_source %in% c("reported process", "ACE table heading") ~
      "Entry or table reported",
    process_classification_source == "ACE event title" ~ "Event-restricted category",
    TRUE ~ "Not reported"
  ))

auctions <- auctions |>
  mutate(entry_rank = if_else(
    program == "COE" & str_detect(rank_key, "^[0-9]+[AB]$"),
    str_remove(rank_key, "[AB]$"), rank_key
  ))

entry_attributes <- competition |>
  select(event_id, program, process_group, entry_rank, entry_key,
         competition_score = score, competition_farm = farm, producer,
         process, process_family, process_classification_source,
         classification_group, competition_variety = variety, region) |>
  group_by(event_id, program, process_group, entry_rank) |>
  filter(n() == 1L) |>
  ungroup()

matched_lots <- auctions |>
  left_join(entry_attributes,
            by = c("event_id", "program", "process_group", "entry_rank")) |>
  mutate(
    score_display = coalesce(competition_score, score),
    farm_display = coalesce(competition_farm, farm),
    variety_display = coalesce(competition_variety, variety),
    process_family = replace_na(process_family, "Not reported"),
    match_status = if_else(is.na(entry_key), "Unmatched or ambiguous", "Unique match")
  )

plot_config <- function(p, filename) plotly::config(
  p, displaylogo = FALSE, responsive = TRUE,
  toImageButtonOptions = list(format = "png", filename = filename)
)

safe_stat <- function(x, fun, min_n = 1L) {
  if (sum(!is.na(x)) < min_n) return(NA_real_)
  fun(x, na.rm = TRUE)
}
```

## V2 overview

::: {.valuebox icon="layers" color="primary"}
### Competition entries
```{r}
cat(if (has_competition) comma(nrow(competition)) else "Awaiting V2 data")
```
:::

::: {.valuebox icon="cup-hot" color="secondary"}
### Auction lots
```{r}
cat(if (has_auctions) comma(nrow(auctions)) else "—")
```
:::

::: {.valuebox icon="currency-dollar" color="success"}
### Reported bids
```{r}
cat(if (has_auctions) comma(sum(!is.na(auctions$final_bid_usd_lb))) else "—")
```
:::

::: {.valuebox icon="calendar3" color="warning"}
### Events
```{r}
cat(comma(n_distinct(c(competition$event_id, auctions$event_id))))
```
:::

::: {.valuebox icon="globe-americas" color="info"}
### Origins
```{r}
cat(comma(n_distinct(c(competition$country, auctions$country))))
```
:::

::: {.valuebox icon="calendar-range" color="dark"}
### Coverage
```{r}
all_years <- c(competition$year, auctions$year)
cat(paste(range(all_years, na.rm = TRUE), collapse = "–"))
```
:::

::: {.card}
### Scope

A **competition entry** is a selected coffee published by a COE/NW competition. An **auction lot** is a sale unit; one entry may be split into 1A/1B lots. Participation measures presence inside this selected specialty institution—not all submissions or national specialty production. Prices extend back to 1999, while named buyers and entry-level process labels have stronger recent coverage. Missing outcomes are never coded as zero.

Score, process and variety are added to a sale lot only when its event/program/process/rank key identifies one unique competition entry. Ambiguous keys remain unmatched rather than duplicating lots.
:::

## Participation

::: {.card}
### Published entries, auction lots and events
```{r}
annual <- bind_rows(
  competition |> count(year, name = "value") |>
    mutate(measure = "Competition entries"),
  competition |> distinct(year, event_id) |> count(year, name = "value") |>
    mutate(measure = "Competition events"),
  auctions |> count(year, name = "value") |>
    mutate(measure = "Auction lots"),
  auctions |> distinct(year, event_id) |> count(year, name = "value") |>
    mutate(measure = "Auction events")
)

plot_ly(annual, x = ~year, y = ~value, color = ~measure,
        type = "scatter", mode = "lines+markers",
        text = ~paste(measure, year, comma(value), sep = "<br>")) |>
  layout(xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Published observations"),
         legend = list(orientation = "h")) |>
  plot_config("annual-participation")
```

Growth can reflect participation, additional country programs, finalist limits, event design or publication coverage. It is an observable first-stage proxy, not yet a national entry rate.
:::

::: {.card}
### Country–year explorer
```{r}
country_year <- competition |>
  group_by(country, year, program) |>
  summarise(events = n_distinct(event_id), entries = n(),
            farms = n_distinct(farm_key[farm_key != "" & !is.na(farm_key)]),
            process_coverage = mean(process_classification_source != "not reported"),
    .groups = "drop")

p <- ggplot(
  country_year,
  aes(
    year, entries, color = program, group = program,
    text = paste(
      country, program, paste0("Entries: ", entries),
      paste0("Events: ", events),
      paste0("Process coverage: ", percent(process_coverage)), sep = "<br>"
    )
  )
) +
  geom_line() +
  geom_point() +
  facet_wrap(vars(country), ncol = 3, scales = "free_y") +
  scale_x_continuous(breaks = scales::breaks_pretty(4)) +
  labs(x = "Year", y = "Published competition entries", color = "Program") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank())

ggplotly(p, tooltip = "text", height = 1450) |>
  plot_config("country-year-participation")
```
:::

## Process specialization

::: {.card}
### Composition with classification provenance
```{r}
process_data <- competition |>
  filter(classification_group == "Entry or table reported") |>
  count(country, year, program, classification_group, process_family,
        name = "entries") |>
  group_by(country, year, program, classification_group) |>
  mutate(share = entries / sum(entries)) |>
  ungroup()

pp <- ggplot(
  process_data,
  aes(
    year, share, color = process_family,
    group = interaction(program, process_family),
    linetype = program,
    text = paste(
      country, program, process_family, paste0("Entries: ", entries),
      percent(share, accuracy = 0.1), sep = "<br>"
    )
  )
) +
  geom_line() +
  geom_point() +
  facet_wrap(vars(country), ncol = 3) +
  scale_x_continuous(breaks = scales::breaks_pretty(4)) +
  scale_y_continuous(labels = percent) +
  labs(x = "Year", y = "Share of reported entries", color = "Process",
       linetype = "Program") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank())

ggplotly(pp, tooltip = "text", height = 1450) |>
  plot_config("process-specialization")
```

This chart uses only **Entry or table reported** observations. **Event-restricted category** remains in the coverage panel because it documents institutional categories, not producer-choice shares from an open multi-process competition.
:::

::: {.card}
### Process information coverage
```{r}
source_year <- competition |> count(year, classification_group, name = "entries") |>
  group_by(year) |> mutate(share = entries / sum(entries)) |> ungroup()
plot_ly(source_year, x = ~year, y = ~share, color = ~classification_group,
        type = "bar", text = ~paste(classification_group,
                                     paste0("Entries: ", entries),
                                     percent(share, accuracy = 0.1), sep = "<br>")) |>
  layout(barmode = "stack", xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Share of entries", tickformat = ".0%"),
         legend = list(orientation = "h")) |>
  plot_config("process-coverage")
```
:::

## Prices and differentiated-market benchmark

::: {.card}
### Score and final bid—selected finalists by country
```{r}
priced <- matched_lots |>
  filter(!is.na(final_bid_usd_lb), !is.na(score_display))

sp <- ggplot(
  priced,
  aes(
    score_display, final_bid_usd_lb, color = process_family, shape = program,
    text = paste(
      country, year, farm_display, process_family, variety_display,
      paste0("Score: ", score_display),
      paste0("USD/lb: ", final_bid_usd_lb), sep = "<br>"
    )
  )
) +
  geom_point(alpha = 0.7) +
  facet_wrap(vars(country), ncol = 3, scales = "free") +
  scale_x_continuous(
    breaks = scales::breaks_pretty(4),
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  scale_y_log10(labels = label_dollar(prefix = "$")) +
  labs(x = "Jury score", y = "Final bid (USD/lb)", color = "Process",
       shape = "Program") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank())

ggplotly(sp, tooltip = "text", height = 1450) |>
  plot_config("score-final-bid")
```

Scores are concentrated around 80–90 because the public ACE tables contain selected finalists rather than all submitted coffees. Each country therefore uses its own observed score range in this figure. The axis is expanded visually, but the underlying scores are unchanged.
:::

::: {.card}
### Auction statistics
```{r}
auctions |> group_by(country, year, program) |>
  summarise(`Auction lots` = n(),
            `Reported bids` = sum(!is.na(final_bid_usd_lb)),
            `Mean bid` = safe_stat(final_bid_usd_lb, mean),
            `Variance` = safe_stat(final_bid_usd_lb, var, 2L),
            `Median bid` = safe_stat(final_bid_usd_lb, median),
            `Minimum` = safe_stat(final_bid_usd_lb, min),
            `Maximum` = safe_stat(final_bid_usd_lb, max), .groups = "drop") |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  datatable(options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE)
```
:::

::: {.card}
### Median auction bid relative to Arabica
```{r}
if (nrow(benchmarks) > 0) {
  arabica <- benchmarks |> filter(series == "Coffee Arabica — World Bank") |>
    group_by(year) |>
    summarise(arabica_usd_lb = mean(value, na.rm = TRUE) / 2.2046226218,
              .groups = "drop")
  ratio <- auctions |> filter(!is.na(final_bid_usd_lb)) |>
    group_by(country, year, program) |>
    summarise(median_bid = median(final_bid_usd_lb), priced_lots = n(),
              .groups = "drop") |>
    left_join(arabica, by = "year") |>
    mutate(ratio = median_bid / arabica_usd_lb)
  rp <- ggplot(
    ratio,
    aes(
      year, ratio, color = program, group = program,
      text = paste(country, program, paste0("Lots: ", priced_lots),
                   paste0(round(ratio, 1), "×"), sep = "<br>")
    )
  ) +
    geom_line() +
    geom_point() +
    facet_wrap(vars(country), ncol = 3, scales = "free_y") +
    scale_x_continuous(breaks = scales::breaks_pretty(4)) +
    labs(x = "Auction year", y = "Median bid / Arabica", color = "Program") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "top", panel.grid.minor = element_blank())

  ggplotly(rp, tooltip = "text", height = 1450) |>
    plot_config("specialty-benchmark-ratio")
}
```

This ratio is a descriptive differentiated-market premium—not a causal effect and not a farm-gate margin.
:::

## Buyer depth

::: {.card}
### Named-buyer coverage
```{r}
buyer_year <- auctions |> group_by(year) |>
  summarise(lots = n(), buyer_reported = sum(!is.na(buyer) & buyer != ""),
            named_buyers = n_distinct(buyer[!is.na(buyer) & buyer != ""]),
            coverage = buyer_reported / lots, .groups = "drop")
plot_ly(buyer_year, x = ~year, y = ~coverage, type = "bar",
        text = ~paste(year, paste0("Named-buyer lots: ", buyer_reported),
                      paste0("Distinct buyers: ", named_buyers),
                      percent(coverage, accuracy = 0.1), sep = "<br>")) |>
  layout(xaxis = list(title = "Year", dtick = 1),
         yaxis = list(title = "Share with named buyer", tickformat = ".0%")) |>
  plot_config("buyer-coverage")
```

Buyer analysis is defensible mainly from 2018 onward. Earlier missing names mean “not published or not recognized,” not “no buyer.”
:::

## Coverage and audit

::: {.card}
### Event-level process coverage
```{r}
datatable(coverage |> mutate(coverage_rate = percent(coverage_rate, accuracy = 0.1)),
          options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE,
          filter = "top")
```
:::

::: {.card}
### Download audit
```{r}
datatable(downloads, options = list(pageLength = 15, scrollX = TRUE),
          rownames = FALSE, filter = "top")
```
:::

::: {.card}
### Raw-table reduction audit
```{r}
datatable(reductions |> arrange(desc(abs(removed))),
          options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE,
          filter = "top")
```

Three fully overlapping sparse summary tables were excluded before harmonization. Remaining reductions still require classification before freezing the ACE analytical base.
:::

::: {.card}
### Interpretation rules

- Entries and auction lots are separate units; split lots never duplicate process choices.
- Process provenance remains explicit: reported, event-restricted or missing.
- Historical prices cover a longer period than buyers and process labels.
- COE and NW remain separately filterable.
- Participation means published finalists, not national adoption.
- Descriptive price ratios and coefficients are not causal effects.
:::

::: {.card}
### Data sources

- [Alliance for Coffee Excellence historical results](https://allianceforcoffeeexcellence.org/competition-auction-results/)
- [World Bank Commodity Price Data](https://www.worldbank.org/en/research/commodity-markets)
- [BLS winery producer price index via FRED](https://fred.stlouisfed.org/series/PCU3121303121300)
:::
